---
title: "Visualization of the 2020 Environmental Performance Index (EPI)"
output:
  html_document:
    toc: true
    toc_float: true
header-includes:
   - \usepackage{svg}
editor_options:
  chunk_output_type: console
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = "hide")
```


# Load libraries
```{r, message = 'hide', results = 'hide'}
library("reshape2")
library("ggplot2")
theme_set(theme_bw(base_size = 12) +
            theme(rect = element_rect(fill = "transparent")))
library("leaflet")
library("rgdal")
library("htmlwidgets")
library("lavaan")
library("lavaanPlot")
```


# Read data
## Clear workspace, set working directory
```{r}
rm(list = ls())
setwd("~/epi")
```


## Load helper functions
These helper functions retrieve, read and process data.
```{r}
source("UtilityFunctions.R")
```


## Import EPI data
Run the helper function `importEPI()` and release the contents of the resulting list into the global environment.
```{r}
epi.list <- importEPI()

list2env(epi.list, .GlobalEnv)
rm(epi.list)
```


## Retrieve world borders
Download the shape file, if it is absent. *Note*: overwrites existing files!
```{r}
if(!file.exists("2020/TM_WORLD_BORDERS-0.3.shp")){
  print("File does not exist, downloading it now:")
  download.file("thematicmapping.org/downloads/TM_WORLD_BORDERS-0.3.zip",
                destfile="world_shape_file.zip")
  system("unzip -u world_shape_file.zip;
         rm world_shape_file.zip Readme.txt")
  }
```


The, read this shape file into a `SpatialPolygonsDataFrame`.
```{r}
world.spdf <- readOGR(dsn = "2020", layer = "TM_WORLD_BORDERS-0.3",
                      verbose = TRUE)
summary(world.spdf)
```


## Collate countries and their names
Unfortunately, as soon as it comes to map the world, geopolitics come to the fore.
This is reflected in the mismatches between country lists due to the different recognition statuses of these countries.

For one, there are distinctly more countries in the shp file:
```{r}
length(world.spdf$NAME) 
length(epi$country)
```


These mismatches stem mainly from different naming schemes.
```{r}
sort(setdiff(world.spdf$NAME, epi$country))
sort(setdiff(epi$country, world.spdf$NAME))
```


Now, collate the naming schemes between the `epi` data.frame and the `world.spdf` SpatialPolygonsDataFrame with the helper function `collateCountryNames()`.
```{r}
epi.world.spdf <- collateCountryNames(epi, world.spdf)
list2env(epi.world.spdf, .GlobalEnv)
rm(epi.world.spdf)
```


Lastly, control if all EPI countries are present in the world map...
```{r}
sort(setdiff(epi$country, world.spdf$NAME))
```
... as the set difference is now equal to zero (i.e., all items in `epi$country` are present also in `world.spdf$NAME`).



Countries without an EPI are miniature states, islands or states currently in an armed conflict (some of the island belong to a nation; e.g., Greenland to Denkmark and Svalbard to Norway. For now, keep it as is and later, these islands could possibly be merged into their mainlands).

```{r}
sort(setdiff(world.spdf$NAME, epi$country))
```



## Add EPI and regions to sp file
```{r}
world.spdf <- merge(world.spdf,
                    epi[, (colnames(epi) %in% c("country", "EPI.new", "region",
                                                "X2019"))], 
                    by.x = "NAME", by.y = "country")
names(world.spdf)
```



## Encode population numbers as millions
```{r}
world.spdf@data$POP2005[which(world.spdf@data$POP2005 == 0)] <- NA
world.spdf@data$POP2005 <- as.integer(world.spdf@data$POP2005) / 1000000 %>%
  round(2)
```


# Choropleth map with leaflet
## Create a color palette for the map
The EPI color scheme:
```{r}
epi.cols <- colorNumeric(palette = "viridis", domain = world.spdf@data$EPI.new,
                         na.color = "gray", reverse = TRUE)
```



The region color scheme:
```{r}
region.cols <- colorFactor(palette = "plasma", domain = world.spdf@data$region,
                           na.color = "gray")
```



## Prepare the text for tooltips
```{r}
tooltips <- paste("Country:", world.spdf@data$NAME, "<br/>",
                  "Region:", world.spdf@data$region, "<br/r>",
                  # "Area:", world.spdf@data$AREA, "<br/>",
                  "Population:", round(world.spdf@data$POP2005, 2), "M",
                  "<br/>",
                  "EPI:", world.spdf@data$EPI.new) %>%
  lapply(htmltools::HTML)
```



## Render map
```{r}
cp.map <- leaflet(world.spdf) %>%
  addTiles() %>%
  setView(lat = 20, lng = 0, zoom = 2.5) %>%
  addPolygons(fillColor = ~ epi.cols(EPI.new), stroke = TRUE, group = "EPI",
              fillOpacity = 0.9, color = "white", weight = 0.5,
              label = tooltips,
              labelOptions = labelOptions(style = list("font-weight" = "normal",
                                                       padding = "3px 8px"),
                                          textsize = "13px",
                                          direction = "auto")) %>%
  addPolygons(fillColor = ~ region.cols(region), stroke = TRUE,
              group = "Region", fillOpacity = 0.5, color = "white",
              weight = 0.5, label = tooltips,
              labelOptions = labelOptions(style = list("font-weight" = "normal",
                                                       padding = "3px 8px"),
                                          textsize = "13px",
                                          direction = "auto")) %>%
  addLegend(pal = epi.cols, values = ~ EPI.new, opacity = 0.9,
            title = "EPI", position = "topleft", group = "EPI") %>%
  addLegend(pal = region.cols, values = ~ region, opacity = 0.5,
            title = "", position = "topleft", group = "Region") %>%
  addLayersControl(overlayGroups = c("EPI", "Region"),
                   options = layersControlOptions(collapsed = FALSE)) %>%
  hideGroup("Region")
cp.map
```



## Export map to html
This takes some time, so only run it if needed.
```{r}
# saveWidget(cp.map, file = "index.html")
```


# Analysis
## Color scale for plotting
```{r}
cols.region <- c("#0D0887", "#5402A3", "#8B0AA5", "#B93289", "#DB5C68",
                 "#F48849", "#FEBC2A", "#F0F921")
```


## National comparison
The 32 performance indicators are displayed by country in boxplots sorted by their median and colored by region.
```{r, fig.width = 15}
ggplot(epi.long[epi.long$Type == "Indicator", ],
       aes(x = reorder(country, EPI.new.value, FUN = median, na.rm = TRUE),
           y = EPI.new.value, fill = region)) +
  geom_boxplot(outlier.shape = 21, outlier.size = 1, alpha = 0.5) +
  scale_fill_manual(values = cols.region) +
  theme(legend.position = "top", legend.title = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text = element_text(size = 5)) +
  ylab("2020 EPIs") +
  xlab("")
# ggsave("EPI_nations.pdf", width = 20, height = 8.27, bg = "transparent")
```

An ordering according to region is apparent. So, let's have a closer look onto these regions.


## Regional comparison
This time, the 32 performance indicators are pooled by region and are displayed in boxplots sorted by their median and colored by region.

```{r}
ggplot(epi.long[epi.long$Type == "EPI", ],
       aes(x = reorder(region, EPI.new.value, FUN = median, na.rm = TRUE),
           y = EPI.new.value, fill = region)) +
  geom_boxplot(outlier.shape = 21, outlier.size = 1, alpha = 0.5) +
  scale_fill_manual(values = cols.region) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 100) +
  ylab("2020 EPI") +
  xlab("")
# ggsave("EPI_regions.pdf", width = 11.29, height = 6.27, bg = "transparent")
```


## EPI vs GDP
```{r}
ggplot(epi.long[epi.long$Type == "EPI", ],
       aes(x = log(X2019 / 1000000), y = EPI.new.value, fill = region)) +
  geom_smooth(aes(group = 1), color = "gray", lty = 2,
              method = "lm", formula = y ~ x, se = FALSE, show.legend = FALSE) +
  geom_point(alpha = 0.5, pch = 21) +
  scale_color_manual(values = cols.region) +
  theme(legend.position = "top", legend.title = element_blank()) +
  ylim(0, 100) +
  ylab("2020 EPI") +
  xlab(expression(paste(log[e], "(2019 GDP)", " [US$]")))
# ggsave("EPI_GDP.pdf", width = 11.29, height = 6.27, bg = "transparent")
```


# Structural equation model
In a structural equation model ...

## Merge datasets
```{r}
epi <- merge(epi,
             world.spdf[, (names(world.spdf) %in% c("ISO3", "AREA",
                                                    "POP2005", "LAT"))],
             by.x = "iso", by.y = "ISO3")
names(epi)
str(epi)
```


## Scale the variables
```{r}
epi[, c(4, 51:54)] <- apply(epi[, c(4, 51:54)], 2, scale)
```


## Specify the model
```{r}
epi.gdp <- 'EPI.new ~ X2019 + POP2005 + LAT
  X2019 ~ LAT + POP2005
  POP2005 ~ AREA'
```


## Run the model and inspect the results
```{r}
fit.epi.gdp <- sem(epi.gdp, data = epi)
summary(fit.epi.gdp, rsq = TRUE)
```


## Inspect modification indices
The modification indices suggest--among others--that there is an unspecified cause that drives GDP and population numbers. In SEM, this unspecified result is modeled by a correlation and encoded with this double tilde `~~`.
```{r}
modificationindices(fit.epi.gdp, minimum.value = 3)
fit.epi.gdp.up <- update(fit.epi.gdp, add = "X2019 ~~ POP2005")
summary(fit.epi.gdp.up, rsq = TRUE)
```


## Visualize the model
```{r}
lavaanPlot(model = fit.epi.gdp.up,
           node_options = list(shape = "box", color = "gray",
                               fontname = "Helvetica"),
           edge_options = list(color = "black"),
           coefs = TRUE, covs = TRUE, stars = c("covs", "regress"),
           labels = list(AREA = "Area", POP2005 = "Population", X2019 = "GDP",
                         LAT = "Latitude", EPInew = "EPI"))
```


